Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-10 paper/paper.tex
Add extension section (WIP)

\begin{lemma}[Adequacy\footnote{\dotv{semtypingopen.v}{adequacy}}]
  \label{lemma:adequacy}
  If $[[InterpR i A S]]$, then $[[S]] \in \CR$.
\end{lemma}

@@

Note that, after this extension, the specification of our logical
relation does not require
any updates. The proof of the fundamental theorem also remains
identical since the complications introduced by $\eta$ are hidden
behind the proofs of the diamond property and the anti-renaming property.
As before, $\ottkw{ne}$ and $\ottkw{nf}$
represent $\beta$-neutral and $\beta$-normal forms, and the
fundamental lemma shows us that every well-typed term has a
$\beta$-normal form. However, in the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}

@@

However, % with a little more effort,
it should be possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}. We leave the proof of the
factorization property and the extraction of a conversion algorithm as
part of our future work.
\fi
\section{Related Work}
\label{sec:relatedwork}

>>>>

\begin{lemma}[Adequacy]
  \label{lemma:adequacy}
  \footnote{\dotv{semtypingopen.v}{adequacy}}
  If $[[InterpR i A S]]$, then $[[S]] \in \CR$.
\end{lemma}

@@

Note that, after this extension, the specification of our logical
relation does not require
any updates. The proof of the fundamental theorem also remains
identical since the complications introduced by $\eta$ are hidden
behind the proofs of the diamond property and the anti-renaming property.
As before, $\ottkw{ne}$ and $\ottkw{nf}$
represent $\beta$-neutral and $\beta$-normal forms, and the
fundamental theorem shows us that every well-typed term has a
$\beta$-normal form. However, in the presence of the $\eta$ reduction
rule, Lemma~\ref{lemma:parnenf} tells us that $\eta$ reduction
preserves $\beta$-normal forms (i.e. does not produce new
$\beta$-redexes). Furthermore, since the $\eta$ reduction rule for
functions strictly decreases the size of the term, the existence of
$\beta\eta$ normal form trivially follows.
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}

@@

However, % with a little more effort,
it should be possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}. We leave the proof of the
factorization property and the extraction of a conversion algorithm as
part of our future work.
\fi

\section{Extensions}
In this section, we discuss how the normalization
proof of \lang can be extended to incorporate features including
natural numbers, dependent pairs, cumulative
universes with subtyping, and $\eta$ law for functions. All these
extensions have been implemented in our development. More advanced
features such as type-directed $\eta$ laws are out of the scope of
this paper, and we instead point the readers to existing work for
techniques that can be readily adapted to our minimal development.

\subsection{Natural Numbers}
\begin{figure}
  \drules[T]{$[[G |- a : A]]$}{Typing for Naturals}{Nat, Zero, Suc, Ind}
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation for Naturals}{Nat}
  \drules[N]{$[[NatOk a]]$}{Natural values}{Zero, Neu, Suc}
\caption{Rules for natural numbers}
\label{fig:nat}
\end{figure}

We can replace the boolean type with the natural number type as our
observable type. By introducing an infinite data type to \lang,
we significantly increase its expressiveness.

The syntactic and semantic rules for natural numbers can be found in
Figure~\ref{fig:nat}. \Rref*{T-Zero, T-Suc} allow us to construct natural
numbers. \Rref*{T-Ind} allows us to eliminate natural numbers
through the induction principle where $[[a]]$ is the base case,
$[[b]]$ the inductive case, and $[[c]]$ the natural number being
eliminated.

\Rref{I-Nat} gives us the semantic
interpretation of natural number; a term $[[a]]$ semantically inhabits
$[[Nat]]$ if $[[a]]$ reduces to some term $[[b]]$ such that $[[NatOk
b]]$ holds.
The $\vdash_\mathbb{N}$ judgment inductively characterizes the
normal form of (potentially open) terms of type $[[Nat]]$ and is
particularly useful in the \rref*{T-Ind} case of the fundamental
theorem. Let $[[rho |= G]]$, the induction hypothesis $[[G |= c :
Nat]]$ tells us that $[[c { rho } =>+ b ]]$ for $[[NatOk b]]$. The proof
then follows by induction over the derivation of $[[NatOk b]]$. When
$[[b]]$ is neutral, we conclude the proof with
\nameref{lemma:adequacy}. When $[[b]]$ is $[[zero]]$, we conclude the
proof with the outer induction hypothesis $[[G |= a : A {zero / x}]]$.
When $[[b]]$ is a successor, we conclude the proof with the outer
induction hypothesis $[[G ++ x : Nat ++ y : A |= b : A { succ x / x
}]]$.

\subsection{Cumulativity}
\label{sec:cumulativity}
The semantic model we build in \cref{sec:logreldep} satisfies
\nameref{lemma:logrelcumulativity}.

\section{Related Work}
\label{sec:relatedwork}
